% ICCV 2025 Paper Template

\documentclass[10pt,twocolumn,letterpaper]{article}

%%%%%%%%% PAPER TYPE  - PLEASE UPDATE FOR FINAL VERSION
\usepackage{iccv}              % To produce the CAMERA-READY version
%\usepackage[review]{iccv}      % To produce the REVIEW version
% \usepackage[pagenumbers]{iccv} % To force page numbers, e.g. for an arXiv version

% Import additional packages in the preamble file, before hyperref
\input{preamble}

\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{amsmath}

\usepackage{epsfig}
\usepackage{multirow}
\usepackage{boldline}
\usepackage{float}     % in the preamble
\usepackage{adjustbox} % already using
\usepackage{booktabs}  % for \toprule etc.
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{stfloats}
\usepackage[accsupp]{axessibility}  % Improves PDF readability for those with disabilities.


% It is strongly recommended to use hyperref, especially for the review version.
% hyperref with option pagebackref eases the reviewers' job.
% Please disable hyperref *only* if you encounter grave issues, 
% e.g. with the file validation for the camera-ready version.
%
% If you comment hyperref and then uncomment it, you should delete *.aux before re-running LaTeX.
% (Or just hit 'q' on the first LaTeX run, let it finish, and you should be clear).
\definecolor{iccvblue}{rgb}{0.21,0.49,0.74}
\usepackage[pagebackref,breaklinks,colorlinks,allcolors=iccvblue]{hyperref}

%%%%%%%%% PAPER ID  - PLEASE UPDATE
\def\paperID{32} % *** Enter the Paper ID here
\def\confName{ICCVW}
\def\confYear{2025}

%%%%%%%%% TITLE - PLEASE UPDATE
\title{MK-UNet: Multi-kernel Lightweight CNN for Medical Image Segmentation}


% Md Mostafijur Rahman
% Radu Marculescu

%%%%%%%%% AUTHORS - PLEASE UPDATE
\author{Md Mostafijur Rahman\\
The University of Texas at Austin\\
Austin, Texas\\
{\tt\small mostafijur.rahman@utexas.edu}
% For a paper whose authors are all at the same institution,
% omit the following lines up until the closing ``}''.
% Additional authors and addresses can be added with ``\and'',
% just like the second author.
% To save space, use either the email address or home page, not both
\and
Radu Marculescu\\
The University of Texas at Austin\\
Austin, Texas\\
{\tt\small radum@utexas.edu}
}

\begin{document}
\maketitle


\begin{abstract}
In this paper, we introduce MK-UNet, a paradigm shift towards ultra-lightweight, multi-kernel U-shaped CNNs tailored for medical image segmentation. Central to MK-UNet is the multi-kernel depth-wise convolution block (MKDC) we design to adeptly process images through multiple kernels, while capturing complex multi-resolution spatial relationships. MK-UNet also emphasizes the images salient features through sophisticated attention mechanisms, including channel, spatial, and grouped gated attention. Our MK-UNet network, with a modest computational footprint of only 0.316M parameters and 0.314G FLOPs, represents not only a remarkably lightweight, but also significantly improved segmentation solution that provides higher accuracy over state-of-the-art (SOTA) methods across six binary medical imaging benchmarks. Specifically, MK-UNet outperforms TransUNet in DICE score with nearly 333$\times$ and 123$\times$ fewer parameters and FLOPs, respectively. Similarly, when compared against UNeXt, MK-UNet exhibits superior segmentation performance, improving the DICE score up to 6.7\% margins while operating with 4.7$\times$ fewer \#Params. Our MK-UNet also outperforms other recent lightweight networks, such as MedT, CMUNeXt, EGE-UNet, and Rolling-UNet, with much lower computational resources. This leap in performance, coupled with drastic computational gains, positions MK-UNet as an unparalleled solution for real-time, high-fidelity medical diagnostics in resource-limited settings, such as point-of-care devices. Our implementation is available at \url{https://github.com/SLDGroup/MK-UNet}.

\end{abstract}

\input{sections/1.introduction}
\input{sections/2.related_work}
\input{sections/2.technique}
\input{sections/3.experiments}
\input{sections/4.ablations}
\input{sections/5.conclusion}

\section*{Acknowledgments}
This work is supported in part by the NSF grant CNS 2007284, and in part by the iMAGiNE Consortium (https://imagine.utexas.edu/).
\vspace{-0.2cm}

{
    \small
    \bibliographystyle{ieeenat_fullname}
    \bibliography{main}
}

\input{sections/6.supply}

\end{document}
